A formal treatment of bidirectional typing

### 2023-10-13 tex/ESOP24/S6-example.tex
[ update ] Add ESOP submission draft
[ tex ] Update Example section

\section{Examples}\label{sec:example}
Our theory works for any language that can be specified by a bidirectional binding signature.
To showcase its general applicability, we present two additional examples:
\begin{enumerate}
\item \PCF with products and let bindings, and
\item spine calculus, a simply typed $\lambda$-calculus with applications in spine form.
\end{enumerate}
The first is a standard exercise in implementing a type synthesiser.
The second example includes not only infinitely many operations but also a type which is not formed by solely one type construct~$\tyOp$.

\subsection{PCF with Products and Let Bindings}\label{subsec:PCF}
Although we have developed a theory for presenting a bidirectional type system and a condition for the decidability of bidirectional type synthesis, designing a bidirectional type system, or its signature, remains a creative activity.\footnote{This has also been the case for parsing: while there are grammars and parser generators, one still has to design a grammar to parse or disambiguate an ambiguous grammar.}
While the focus of this paper is not to design bidirectional type systems, we simply adopt the example of a bidirectional \PCF from \citet{Wadler2022}, which extends bidirectional simply typed $\lambda$-calculus in \Cref{sec:key-ideas} by adding the following rules:
\bgroup
\small
  \begin{mathpar}
    \inferrule{ }{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}
    \and
    \inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}
    \and
    \inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}
    \and
    \inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}
    \and
    \inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}
    \\
    \inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, t \chk A}
    \and
    \inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}
  \end{mathpar}
\egroup

%In this language, there are additional type constructs other than $\bto$, so first we extend $\Sigma_{\bto}$ (\Cref{ex:type-signature-for-function-type}) with operations
In this language, we introduce additional type constructs apart from $\bto$.
To do so, we extend $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with the operations $\mathsf{nat}$ and $\mathsf{times}$ such that
\[
  \arity({\mathsf{nat}}) = 0 \quad\text{and}\quad \arity({\mathsf{times}}) = 2
\]
introducing $\mathtt{nat}$ and $A \times B$ as $\tyOp_{\mathsf{nat}}$ and $\tyOp_{\mathsf{times}}(A, B)$, respectively.
Following this, we also extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with an operation for each rule above:
\begin{flalign*}
  &  \aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}
  && \aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}} \\
  &  \aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}
  && \aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}
  && \aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\syn}} \\
  &  \aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}
  && \aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}
\end{flalign*}
With these operations, a straightforward check shows that every rule is mode-correct, therefore this is a mode-correct bidirectional type system.
As a result, we can instantiate the type synthesiser presented in \Cref{sec:type-synthesis} for this system, eliminating the need for re-implementation.

\subsection{Spine Calculus}\label{subsec:spine}
A spine application, denoted as $t\;u_1\;\ldots\;u_n$, is a variant of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
This arrangement allows direct access to the head term, making it practical in various applications.
For instance, \Agda's core language employs this form of application, as does its reflected syntax used for metaprogramming.

@@

To illustrate this, let us consider the following rule:
\bgroup
\small
  \begin{mathpar}
    \inferrule{\Gamma \vdash t \syn A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right) \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
  \end{mathpar}
\egroup
We use the same type signature $\Sigma_{\bto}$ but extend the bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
\[
  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right)^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
\]
for each $n : \N$.
Then each spine application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.

>>>>

\section{Examples}\label{sec:example}
In this section, we use two more examples to exhibit the applicability of our approach.
This first one has infinitely many operations and the second one includes many more constructs than simply typed $\lambda$-calculus with some common features in a programming language.

\subsection{Spine application}\label{subsec:spine}
A spine application $t\;u_1\;\ldots\;u_n$ is a variant form of application that consists of a head term $t$ and an indeterminate number of arguments $u_1\;u_2\;\dots\;u_n$.
This arrangement allows direct access to the head term, making it practical in various applications.
For instance, \Agda's core language employs this form of application, as does its reflected syntax used for metaprogramming.

@@

To illustrate this, let us consider the following rule:
\bgroup
\small
  \begin{mathpar}
    \inferrule{\Gamma \vdash t \syn A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right) \and \Gamma \vdash u_1 \chk A_1 \and \dots \and \Gamma \vdash u_n \chk A_n}{\Gamma \vdash t\;u_1\;\ldots\;u_n \syn B}
  \end{mathpar}
\egroup
We use the same type signature $\Sigma_{\bto}$ but extend the signature $\Omega_{\Lambda}^{\Leftrightarrow}$ with 
\[
  \aritysymbol{\mathsf{app}_n}{A_1,\ldots, A_n, B}{A_1 \bto \left(A_2 \bto \left(\dots \bto \left(A_n \bto B\right)\ldots\right)\right)^{\syn}, A_1^{\chk}, \ldots, A_n^{\chk}}{B}
\]
for each $n : \N$, so each application $t\;u_1\;\ldots\;u_n$ can be introduced as $\tmOp_{\mathsf{app}_n}(t; u_1; \ldots; u_n)$.
\subsection{Computational calculi}\label{subsec:PCF}
The code size of a type synthesiser grows proportionally to the number of type constructs and the number of arguments for each language construct, making it a tedious task to implement.
Even for a simply typed language which is generally straightforward to implement a synthesiser individually, having a generator can reduce the effort significantly.
For illustrative purposes, let us consider the Moggi's computational $\lambda$-calculus~\cite{Moggi1989} with additional constructs.

We first introduce additional type constructs apart from $\bto$ by extending $\Sigma_{\bto}$ (as shown in \Cref{ex:type-signature-for-function-type}) with $\mathsf{nat}$, $\mathsf{prod}$, $\mathsf{sum}$, and $\mathsf{T}$ such that
\[
  \arity({\mathsf{nat}}) = 0\qquad
  \arity({\mathsf{T}}) = 1\qquad
  \arity({\mathsf{prod}}) = \arity({\mathsf{sum}}) = 2
\]
introducing $\mathtt{nat}$, $A \times B$, $A + B$, $T(A)$ as $\tyOp_{\mathsf{nat}}$, $\tyOp_{\mathsf{prod}}(A, B)$, $\tyOp_{\mathsf{sum}}(A, B)$, and $\tyOp_{\mathsf{T}}(A)$ respectively.
The bidirectional binding signature $\Omega_{\Lambda}^{\Leftrightarrow}$ is then extended with operations in following table.

%\begin{table}
\bgroup
  \renewcommand{\arraystretch}{2}
  \centering\small
\begin{tabular}{c | l}
  Rules & Operations \\ \hline\hline
  $\inferrule{}{\Gamma \vdash \mathtt{z} \chk \mathtt{nat}}$ &  $\aritysymbol{\mathsf{z}}{\cdot}{\cdot}{\mathtt{nat}^{\chk}}$ 
  \\
 $\inferrule{\Gamma \vdash t \chk \mathtt{nat}}{\Gamma \vdash \mathtt{s}(t) \chk \mathtt{nat}}$ & $\aritysymbol{\mathsf{s}}{\cdot}{\mathtt{nat}^{\chk}}{\mathtt{nat}^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \syn \mathtt{nat} \and \Gamma \vdash t_0 \chk A \and \Gamma, x : \mathtt{nat} \vdash t_1 \chk A}{\Gamma \vdash \mathtt{ifz}(t_0; x.t_1)(t) \chk A}$ & $\aritysymbol{\mathsf{ifz}}{A}{\mathtt{nat}^{\syn}, A^{\chk}, A^{\chk}}{A^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \chk A \and \Gamma \vdash u \chk B}{\Gamma \vdash (t, u) \chk A \times B}$ & $\aritysymbol{\mathsf{pair}}{A, B}{A^{\chk}, B^{\chk}}{A \times B^{\chk}}$  \\
 $\inferrule{\Gamma \vdash t \syn A \times B}{\Gamma \vdash \mathtt{proj}_1(t) \syn A}$ & $\aritysymbol{\mathsf{proj}_1}{A, B}{A \times B^{\syn}}{A^{\syn}}$ \\

 $\inferrule{\Gamma \vdash u \syn A \times B}{\Gamma \vdash \mathtt{proj}_2(u) \syn A}$ & $\aritysymbol{\mathsf{proj}_2}{A, B}{A \times B^{\syn}}{B^{\chk}}$ \\

 $\inferrule{\Gamma \vdash t \chk A}{\Gamma \vdash \mathtt{inj}_1(t) \chk A + B}$ & $\aritysymbol{\mathsf{inj}_1}{A, B}{A^{\chk}}{A + B^{\chk}}$ \\

 $\inferrule{\Gamma \vdash u \chk B}{\Gamma \vdash \mathtt{inj}_2(u) \chk A + B}$ & $\aritysymbol{\mathsf{inj}_2}{A, B}{B^{\chk}}{A + B^{\chk}}$ \\

 $\inferrule{\Gamma \vdash u \syn A + B \and \Gamma, x_1 : A \vdash t_1 \chk C \and \Gamma, x_2 : B \vdash t_2 \chk C}{\Gamma \vdash \mathtt{case}(u; x_1. t_1; x_2. t_2) \chk C}$ & $\aritysymbol{\mathsf{case}}{A, B, C}{{A + B}^{\syn}, [A]C^{\chk}, [B]C^{\chk}}{C^{\syn}}$ \\

 $\inferrule{\Gamma, x : A \vdash t \chk A}{\Gamma \vdash \mu x.\, t \chk A}$ & $\aritysymbol{\mathsf{mu}}{A}{[A]A^{\chk}}{A^{\chk}}$ \\

 $\inferrule{\Gamma \vdash t \syn A \and \Gamma, x : A \vdash u \chk B}{\Gamma \vdash \mathtt{let}\;x = t\;\mathtt{in}\;u\chk B}$ & 
  $\aritysymbol{\mathsf{let}}{A, B}{A^{\syn},[A]B^{\chk}}{B^{\chk}}$ \\
 $\inferrule{\Gamma \vdash t \syn A}{\Gamma \vdash \mathtt{ret}(t) \syn T(A)}$ & $\aritysymbol{\mathsf{ret}}{A}{A^{\syn}}{T(A)^{\syn}}$ \\
 $\inferrule{\Gamma \vdash t \syn T(A) \and \Gamma, x : A \vdash u \syn T(B)}{\Gamma \vdash \mathtt{bind}(t; x.u) \syn T(B)}$ & $\aritysymbol{\mathsf{bind}}{A, B}{T(A)^{\syn}, [A]T(B)^{\syn}}{T(B)^{\syn}}$
\end{tabular}
\egroup

The extended language has `only' 15 constructs, far fewer than what the surface language of a programming language in use would have. 
However, if we implement a proof-relevant type synthesiser following \citet{Wadler2022}, then there will be nearly 100 cases to check on top of other auxiliary definitions.

### 2023-10-13 tex/ESOP24/appendix.tex
[ tex ] Update Example section

\begin{definition}\label{def:syn-var}
  For an argument list $\biargs$, the set of \emph{synthesised type variables} is defined inductively by 
  \begin{align*}
    \synvar(\cdot)                   & = \emptyset  \\
    \synvar(\biargs, \chkbiarg[n+1]) & = \phantom{\fv(A_{n+1}) \cup {}} \synvar(\biargs) \\
    \synvar(\biargs, \synbiarg[n+1]) & = \fv(A_{n+1}) \cup           \synvar(\biargs).
  \end{align*}
\end{definition}

>>>>

\begin{definition}\label{def:syn-var}
  The subset $\fv(\Delta)$ of variables in an extension context~$\Delta$ is defined by\/ $\fv(\cdot) = \emptyset$ and\/ $\fv(\Delta, A) = \fv(\Delta) \cup \fv(A)$.
  For an argument list $\biargs$, the set of \emph{synthesised type variables} is defined by 
  \begin{align*}
    \synvar(\cdot)                   & = \emptyset  \\
    \synvar(\biargs, \chkbiarg[n+1]) & = \phantom{\fv(A_{n+1}) \cup {}} \synvar(\biargs) \\
    \synvar(\biargs, \synbiarg[n+1]) & = \fv(A_{n+1}) \cup           \synvar(\biargs).
  \end{align*}
\end{definition}
